---
title: "Practical 1 - Medical Databases: Predicting Sepsis from ICU Data"
author: "Heather MacTavish"
date: "2026-05-23"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5,
  dpi        = 110,
  cache      = TRUE,
  cache.lazy = FALSE
)

GLOBAL_SEED <- 42
set.seed(GLOBAL_SEED)

options(
  digits     = 4,
  scipen     = 999,
  dplyr.summarise.inform = FALSE
)
```

> **Submission.** Knit your Rmd file into HTML and submit it via Brightspace. Include a link to your git repository (GitHub or git.cs.dal.ca) as a comment on your submission.

---

## 0 Setup

```{r packages}
library(tidyverse)    # dplyr, ggplot2, tidyr, purrr, ...
library(tidymodels)   # rsample, recipes, parsnip, workflows, tune, yardstick, ...
library(ranger)       # random forest engine
library(xgboost)      # gradient-boosted trees engine
library(glmnet)       # penalised logistic regression engine
library(here)
library(fs)

theme_set(theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank()))
```

---

## 1 Introduction

[Sepsis](https://www.who.int/news-room/fact-sheets/detail/sepsis) is life-threatening organ dysfunction caused by a dysregulated immune response to infection.
It accounts for roughly 20 % of global deaths, and each hour of delayed treatment increases mortality by 4–8 %.

The [PhysioNet/CinC Challenge 2019](https://physionet.org/content/challenge-2019/1.0.0/) dataset contains hourly ICU records for ~40 000 patients from two US teaching hospitals.
Each record (row) has 8 vital signs, 26 laboratory measurements, 6 demographic fields, and a binary `SepsisLabel` (set to 1 up to 6 hours *before* clinical onset so that early-warning models can be trained).
We simplify the temporal problem to a static one by taking one random hourly record per patient, giving a standard tabular classification task with realistic clinical challenges: class imbalance, high missingness, and mixed data types.

### Question 1

**Q1. (2 marks)** Look up the [PhysioNet variable descriptions](https://physionet.org/content/challenge-2019/1.0.0/) and include an explanation of the dataset into this notebook (one row per variable: name, description, units).  


```{r}
variable_table <- data.frame(
  Variable = c("HR", "O2Sat", "Temp", "SBP", "MAP", "DBP", "Resp", "EtCO2",
               "BaseExcess", "HCO3", "FiO2", "pH", "PaCO2", "SaO2", "AST",
               "BUN", "Alkalinephos", "Calcium", "Chloride", "Creatinine",
               "Bilirubin_direct", "Glucose", "Lactate", "Magnesium",
               "Phosphate", "Potassium", "Bilirubin_total", "TroponinI",
               "Hct", "Hgb", "PTT", "WBC", "Fibrinogen", "Platelets",
               "Age", "Gender", "Unit1", "Unit2", "HospAdmTime",
               "ICULOS", "SepsisLabel"),
  Description = c(
    "Heart rate", "Pulse oximetry", "Temperature",
    "Systolic blood pressure", "Mean arterial pressure",
    "Diastolic blood pressure", "Respiration rate",
    "End tidal carbon dioxide", "Measure of excess bicarbonate",
    "Bicarbonate", "Fraction of inspired oxygen", "pH",
    "Partial pressure of carbon dioxide from arterial blood",
    "Oxygen saturation from arterial blood",
    "Aspartate transaminase", "Blood urea nitrogen",
    "Alkaline phosphatase", "Calcium", "Chloride",
    "Creatinine", "Bilirubin direct", "Serum glucose",
    "Lactic acid", "Magnesium", "Phosphate", "Potassium",
    "Total bilirubin", "Troponin I", "Hematocrit",
    "Hemoglobin", "Partial thromboplastin time",
    "Leukocyte count", "Fibrinogen", "Platelets",
    "Age (100 for patients 90 or above)", "Gender",
    "Administrative identifier for ICU unit (MICU)",
    "Administrative identifier for ICU unit (SICU)",
    "Hours between hospital admission and ICU admission",
    "ICU length of stay since ICU admission",
    "Sepsis outcome label"
  ),
  Units = c(
    "beats/min", "%", "Deg C", "mm Hg", "mm Hg", "mm Hg",
    "breaths/min", "mm Hg", "mmol/L", "mmol/L", "%",
    "N/A", "mm Hg", "%", "IU/L", "mg/dL", "IU/L",
    "mg/dL", "mmol/L", "mg/dL", "mg/dL", "mg/dL",
    "mg/dL", "mmol/dL", "mg/dL", "mmol/L", "mg/dL",
    "ng/mL", "%", "g/dL", "seconds", "count*10^3/µL",
    "mg/dL", "count*10^3/µL", "years", "Female (0) or Male (1)",
    "N/A", "N/A", "hours", "hours", " For sepsis patients, SepsisLabel is 1 if t≥tsepsis−6 and 0 if t<tsepsis−6. For non-sepsis patients, SepsisLabel is 0."
  )
)

knitr::kable(variable_table)
```

### Question 2

**Q2. (2 marks)** Then choose **three variables** you hypothesise might be strong predictors of sepsis and write one sentence each explaining your reasoning.  There is no right or wrong answer to this, but you should think about your and problem before diving into the EDA and prediction!
*(Hint: look up how sepsis is clinically diagnosed.)*  

1. *Respiratory Rate (Resp)*

I hypothesize that respiratory rate will be a strong predictor of sepsis because an elevated respiratory rate is included in both the SIRS (Systemic Inflammatory Response Syndrome) and qSOFA (Quick Sequential Organ Failure Assessment) criteria and is often one of the earliest signs of systemic infection.

2. *Systolic Blood Pressure (SBP)*

I hypothesize that systolic blood pressure will be a strong predictor of sepsis because reduced blood pressure is commonly associated with sepsis-induced circulatory dysfunction and is included in the qSOFA criteria.

3. *Temperature (Temp)*

I hypothesize that temperature will be a strong predictor of sepsis because fever (>38°C) or hypothermia (<36°C) are common physiological responses to infection and are included in the SIRS criteria used to identify patients at risk of sepsis.


Citation for how sepsis is clinically diagnosed:
Singer M, Deutschman CS, Seymour CW, Shankar-Hari M, Annane D, Bauer M, et al. The Third International Consensus Definitions for Sepsis and Septic Shock (Sepsis-3). JAMA. 2016 Feb 23;315(8):801–10. doi:[10.1001/jama.2016.0287](https://jamanetwork.com/journals/jama/fullarticle/2492881)

---

## 2 Data Loading and Splitting

To avoid inadvertent data leakage from your test set, we are going to perform our test-train split before even our EDA and cleaning related analyses. This is not always done but is a good practice if you aren't using an external/independent test dataset.

### 2.1 Load the data

I've already done the random sampling of the dataset to remove the time-series elements but the original dataset can be acquired directly from [PhysioNet/CinC Challenge 2019](https://physionet.org/content/challenge-2019/1.0.0/) 
```{r data-paths}
data_dir  <- here("data")
cache_dir <- fs::path(data_dir, "cache")
fs::dir_create(cache_dir)

data_url <- "http://maguire-lab.github.io/health_data_science_research_2026/static_files/practicals/lab1_data/physionet_subsample_modified_1.0.0.rds"
rds_path <- fs::path(cache_dir, "physionet_subsample_1.0.0.rds")
```

Rmd lets you cache expensive or slow operations (like downloading big files or complicated processing) so they only re-run if the code and variables have changed.  This can be very useful although can sometimes cause some inadvertent issues that reduce reproducibility!  You should also do a completely clean re-run of everything once your analyses are finalised if possible.

```{r load-data, cache = TRUE}
if (!file.exists(rds_path)) {
  download.file(data_url, rds_path, mode = "wb", quiet = TRUE)
}
sepsis <- readRDS(rds_path)
```

### 2.2 Train / test split

We split the data **before** any exploration so that analysis choices cannot be influenced by the test set - this is a key discipline in applied machine learning.  Stratified splitting preserves the sepsis rate in both halves.

```{r split}
# will make later things easier to do!s
sepsis <- sepsis |>
  mutate(SepsisLabel = factor(SepsisLabel,
                              levels = c(1, 0),
                              labels = c("Sepsis", 
                                         "No Sepsis")))

# splitting
split <- initial_split(sepsis, prop = 0.75, strata = SepsisLabel)

train <- training(split)
test  <- testing(split)

cat("Training rows:", nrow(train), "\n", "\n")
cat("Test rows:    ", nrow(test),  "\n")
```

### Question 3

**Q3. (1 mark)** What percentage of our dataset is being assigned to test?

25% of the dataset is assigned to test, since `prop = 0.75` means that 75% goes to training and the remaining 25% goes to the test set.

--- 

## 3 Exploratory Data Analysis

Again, do your EDA on `train` only - we do not look at `test` until Section 6.

### 3.1 Class balance

### Question 4

**Q4. (2 marks).** Uncomment and complete the code below to check the class balance in the **training** set.
`count()` tallies rows per group; add a `prop` column to show each class as a fraction of the total.

```{r q4-class-balance}
train |>
 count(SepsisLabel) |>
 mutate(prop = n / sum(n))
```

### Question 5

**Q5. (2 mark)** Using your solution to Q4, complete this plotting code to create a barplot of the relative sepsis label balance.  Change the colour of the sepsis bar to green.

```{r class-balance-plot, fig.height = 4}
train |>
 count(SepsisLabel) |>
 mutate(prop = n / sum(n)) |>
  ggplot(aes(SepsisLabel, 
             prop, 
             fill = SepsisLabel)) +
  geom_col() +
  geom_text(aes(
    label = scales::percent(prop, 
                            accuracy = 0.01)), 
    vjust = -0.4) +
 scale_y_continuous(labels = scales::percent) +
 scale_fill_manual(values = c("No Sepsis" = "grey60",
                              "Sepsis"    = "green")) +
 labs(x = NULL, 
      y = "Percentage of Trainining Data", 
      title = "Sepsis Class Balance") +
 theme(legend.position = "none")
```

### Question 6

**Q6. (3 marks)**  Given the class imbalance, which of the following 2 metrics might be the most misleading and why? accuracy, precision, recall, F1-score, ROC-AUC, PR-AUC?

Accuracy and ROC-AUC are the two most misleading metrics given the class imbalance. Accuracy is misleading because sepsis is so rare in the dataset that if it just predicted "No Sepsis" for every patient it would still get 96.8% accuracy without actually identifying a single sepsis case. ROC-AUC can also be misleading because it accounts for true negatives in the denominator, and there are so many non-sepsis patients that the model can still look like its performing well overall even if it's actually doing a poor job of identifying the sepsis cases.

PR-AUC would be more appropriate than the ROC-AUC in this case because it focuses specifically on how well the model can identify the rare positive class.

### 3.2 Missingness

```{r missingness-bar, fig.height = 5}
train |>
  summarise(across(everything(), ~ mean(is.na(.x)))) |>
  pivot_longer(everything(),
               names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  ggplot(aes(pct_missing, fct_reorder(variable, pct_missing))) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Missing data by variable (training set)",
       x = "Percent missing", y = NULL)
```

### Question 7

**Q7. (1 mark)**
Modify the above to plot only missingness for Sepsis positive data
```{r missingness-sepsis-bar, fig.height = 5}
train |>
  filter(SepsisLabel == "Sepsis") |>
  summarise(across(everything(), 
                   ~ mean(is.na(.x)))) |>
  pivot_longer(everything(),
               names_to = "variable", 
               values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  ggplot(aes(pct_missing, 
             fct_reorder(variable, 
                         pct_missing))) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = scales::percent_format(
    accuracy = 1)) +
  labs(title = "Missing data by variable (training set) for only sepsis patients",
       x = "Percent missing", y = NULL)
```

### Question 8

**Q8. (3 marks)** Which three variables have the highest missingness? In an ICU context, why might a lab value be missing - is this likely to be random, or informative? Does this mean you should drop or impute these values?

!!!!!!

The three variables with the highest missingness are:

1. TroponinI
2. Fibrinogen
3. Bilirubin_direct

In an ICU context, missing lab values are typically not random and are usually informative, because clinicians only order certain tests when they suspect a specific condition (e.g., liver dysfunction or cardiac injury). This means that missingness can reflect clinical decision-making instead of data collection failure.

I don't think that the values should be imputed because it is informative, but there have been studies done that suggest that another variable could be included in the prediction model that indicates it the variable has missingness or not, implying that for example, when looking at cholesterol, the lower the cholesterol measurement, the more favourable, but the most favourable measurement is the lack of one because that would imply that the doctor was not worried about the outcomes that that variable is typically used to detect (Groenwold, 2020).

Groenwold RHH. Informative missingness in electronic health record systems: the curse of knowing. Diagn Progn Res. 2020 Jul 2;4:8. doi: [10.1186/s41512-020-00077-0](https://doi.org/10.1186/s41512-020-00077-0). PMID: 32699824; PMCID: PMC7371469.

### 3.3 Vital-sign distributions by outcome

```{r vitals-density, fig.height = 5}
vitals <- c("HR", "O2Sat", "Temp", "SBP", "MAP", "DBP", "Resp")

train |>
  select(all_of(vitals), SepsisLabel) |>
  pivot_longer(-SepsisLabel) |>
  filter(!is.na(value)) |>
  ggplot(aes(value, fill = SepsisLabel)) +
  geom_density(alpha = 0.55, colour = NA) +
  facet_wrap(~ name, scales = "free", ncol = 4) +
  scale_fill_manual(values = c("No Sepsis" = "grey60",
                               "Sepsis"    = "firebrick")) +
  labs(x = NULL, y = "Density", fill = NULL,
       title = "Vital-sign distributions by sepsis status")
```

### Question 9

**Q9 (2 marks).** Modify the code above to plot the distribution of **three other lab variables** of your choice. 

```{r q9-lab-density}
# I'm addig WBC, Glucose, and Creatinine
labs <- c("WBC", "Glucose", "Creatinine")

train |>
  select(all_of(labs), SepsisLabel) |>
  pivot_longer(-SepsisLabel) |>
  filter(!is.na(value)) |>
  ggplot(aes(value, fill = SepsisLabel)) +
  geom_density(alpha = 0.55, colour = NA) +
  facet_wrap(~ name, scales = "free", ncol = 4) +
  scale_fill_manual(values = c("No Sepsis" = "grey60",
                               "Sepsis"    = "firebrick")) +
  labs(x = NULL, y = "Density", fill = NULL,
       title = "Laboratory value distributions by sepsis status")

```

Correlation plot of all variables by similarity of their correlation in Sepsis and Non-Sepsis

```{r}
cor_mat <- function(df) {
  df |>
    select(where(is.numeric)) |>           
    cor(use = "pairwise.complete.obs")     
}

m_ns  <- cor_mat(train |> filter(SepsisLabel == "No Sepsis"))
m_sep <- cor_mat(train |> filter(SepsisLabel == "Sepsis"))
m_dif <- m_sep - m_ns

m_dif[is.na(m_dif)] <- 0
hc <- hclust(as.dist(1 - abs(m_dif)), method = "average")
var_order <- hc$labels[hc$order]

m_dif |>
  as.data.frame() |>
  tibble::rownames_to_column("var1") |>
  tidyr::pivot_longer(-var1, names_to = "var2", values_to = "diff") |>
  mutate(var1 = factor(var1, levels = var_order),
         var2 = factor(var2, levels = var_order)) |>
  ggplot(aes(var1, var2, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-0.5, 0.5),
                       oob = scales::squish, name = "Δ r") +
  coord_fixed() +
  labs(x = NULL, y = NULL,
       title = "Change in pairwise correlations: Sepsis − No Sepsis") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid  = element_blank())
```

### 3.4 Dimensionality reduction

PCA compresses all numeric features into a small number of axes and lets us ask: *do septic and non-septic patients separate at all in the joint feature space?*
We build it inside a small `recipe` so the preprocessing (imputation, normalisation) is explicit and reproducible.

Recipes are a special type of function in `tidymodels` that differ from normal R functions in key ways (internal state).

A plain R function executes immediately and is stateless i.e., every call recomputes from scratch.
```
rscale_hr <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)

train$HR_z <- scale_hr(train$HR)
test$HR_z  <- scale_hr(test$HR)   # BUG: uses test set's own mean/sd
```

The test set is standardised using statistics computed *from the test set*, which leaks information and means train and test are now on different scales. To do it correctly you'd have to manually save `mean(trainHR)` and `sd(trainHR)` and `sd(trainHR)` and `sd(trainHR)` and reuse them and then do the same for every other transformation, for every column, for imputation, encoding, etc. It gets unmanageable fast.

A recipe on the otherhand is a two-phase object: first you declare the steps, then you `prep()` it on training data (which estimates and stores any required parameters means, sds, imputation values, factor levels, PCA loadings…), then you `bake()` it on any new data to apply the same learned transformations.

```{r pca-prep}
pca_prep <- recipe(SepsisLabel ~ ., data = train) |>
  step_rm(ICULOS, HospAdmTime) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 10, id = "pca") |>
  prep()

pca_scores <- bake(pca_prep, new_data = NULL)
```

```{r pca-scatter, fig.width = 5.5, fig.height = 4}
pca_scores |>
  ggplot(aes(PC01, PC02, colour = SepsisLabel)) +
  geom_point(alpha = 0.75, size = 0.7) +
  scale_colour_manual(values = c("No Sepsis" = "grey60",
                                 "Sepsis"    = "firebrick")) +
  labs(title = "PCA: PC1 vs PC2", colour = NULL) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2)))
```

```{r pca-scree, fig.height = 3}
tidy(pca_prep, id = "pca", type = "variance") |>
  filter(terms == "percent variance") |>
  ggplot(aes(component, value)) +
  geom_col(fill = "steelblue") +
  labs(x = "Principal component", y = "% variance explained",
       title = "PCA scree plot")
```


The two classes overlap heavily in the PC1/PC2 projection — this is common in clinical data where no single linear combination cleanly separates cases.

### Question 10

**Q10 (2 marks).** What does this tell us about the potential decision boundary our predictive model would have to fit?  Which models easily support this type of boundary?

The heavy overlap between sepsis and non-sepsis cases in the PCA plot suggests that the two groups cannot be easily separated using a simple linear boundary. This indicates that the relationship between the predictors and sepsis is likely non-linear and more complex. Models such as random forests or XGBoost may be better at capturing these patterns because they use multiple decision trees rather than relying on a single linear combination of predictors.

---

## 4 Feature Engineering

Let's create a few new features likely to be informative for sepsis:

### Question 11

**Q11. (1 mark)** Modify the following code to also add an additional feature called `shock_index` which is defined as heart rate divided by the systolic blood pressure. 

```{r feature-engineering}
add_features <- function(df) {
  df |>
    mutate(
      pulse_pressure  = SBP - DBP,
      bun_creat_ratio = BUN / Creatinine,  # > 20 suggests pre-renal AKI
      age_group = cut(Age,
                      breaks = c(-1, 40, 65, 
                                 80, 120),
                      labels = c("<=40", "41-65", 
                                 "66-80", "80+")),
      shock_index = HR / SBP # added shock_index
    )
}
train <- add_features(train)
test  <- add_features(test)
```

## 5 Modelling

Now lets try to fit some predictive models to the training set

### 5.1 Cross-validation folds

Five-fold stratified CV on the training set, used to select hyperparameters (see below) by repeatedly training the model with different hyperparameter values on 4/5ths of the data and evaluating on each of the held-out 1/5ths.s   

```{r cv-folds}
folds <- vfold_cv(train, v = 5, strata = SepsisLabel)
```

### 5.2 Preprocessing recipes

Not all models can handle the same type of data and some models (like linear models) benefit from normalising values to similar ranges. Most models will ignore missing values but we are going to add a missingness-indicator columns so the model can learn from *which* values are absent.

We are going to use `tidymodel` recipes again so we can apply the training normalisation etc to the test set to prevent leakage.

```{r recipes}
base_recipe <- recipe(SepsisLabel ~ ., data = train) |>
  step_rm(ICULOS, HospAdmTime) |>
  step_indicate_na(all_numeric_predictors(), prefix = "miss") |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>
  step_zv(all_predictors())

glm_recipe <- base_recipe |>
  step_normalize(all_numeric_predictors())
```

### 5.3 Model specifications

Now let's specify our 3 models:
- Logistic regression: a linear baseline. Outputs P(Sepsis∣X) via a logistic link function (i.e., squeezing our linear model through a sigmoid function that constrains between 0 and 1). By default `glmnet` uses a regularised model.  This means we are penalising the model complexity (the size of the weights/coefficients are added to the cost function).  Specifically, we are using an cross-validation optimied combination of L1 and L2 penalties (absolute and squared weight/coefficient sizes).  This is useful as many of our ~40 lab variables are correlated or weakly informative.

- Random forest: an ensemble of decision trees, each fit on a bootstrap sample with a random subset of features at each split. Captures non-linearities and interactions automatically, and handles mixed-scale features without preprocessing. Robust to outliers and (via `ranger`) to NAs.

- XGBoost: gradient-boosted trees. Builds trees sequentially, each one correcting the residual errors of the previous ensemble. Typically the strongest tabular performer, but the most sensitive to model (hyper)parameters and easy to overfit without tuning.

Each of these has several hyperparameters (parts of the model we don't directly learn from the data during fitting; instead we choose them before fitting and select the best values by cross-validation). Marking a hyperparameter with `tune()` is a placeholder that says "leave this blank for now — we'll fill it in during the tuning step." The values will later be chosen by tune_grid() based on cross-validated performance on the training folds.

### Question 12

**Q12 4 marks**: Using R's documentation (e.g., `?logistic_reg`) look up and briefly explain each of the hyperparamters we are tuning in cross-validation for each of the 3 models:s

```{r model-spec}
glm_spec <- logistic_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("classification")

# hyperparameters: penalty scaling factor and proportion L1 vs L2s
rf_spec <- rand_forest(trees = 500, min_n = tune()) |>
  set_engine("ranger", importance = "permutation", num.threads = 1) |>
  set_mode("classification")

xgb_spec <- boost_tree(
  trees = 500, tree_depth = tune(), learn_rate = tune(),
  loss_reduction = tune(), sample_size = tune(),
   min_n = tune()
) |>
  set_engine("xgboost", nthread = 1) |>
  set_mode("classification")
```



!!!!!! 

*Logistic Regression*
- **penalty**: Controls the strength of regularization. *Larger values shrink coefficients more strongly.*

- **mixture**: Controls the type of regularization used. A value of 0 corresponds to ridge regression (L2 penalty), 1 corresponds to lasso regression (L1 penalty), and values between 0 and 1 produce elastic net regularization.

*Random Forest*:
- **min_n**: The minimum number of data points in a node that are required for the node to be split further.

*XGBoost*:
- **tree_depth**: The maximum depth of each decision tree.
- **learn_rate**: Controls how much each new tree contributes to the model and how much it corrects the errors of previous trees.
- **loss_reduction**: The minimum reduction in the loss function required before an additional split is made. Larger values make the model more conservative by preventing unnecessary splits.
- **sample_size**: The proportion of training observations randomly sampled for each boosting iteration. Sampling can reduce overfitting and improve model robustness.
- **min_n**: The minimum number of data points in a node that are required for the node to be split further.


!!!!!!!!!!!!!!!! more detail required re-write them !!!!!!!!!


### 5.4 Tune all three models

Now let's specify our workflow to use cross-validation to tune all models and compare across them.

```{r workflow-set}
wf_set <- workflow_set(
  preproc = list(glm = glm_recipe, rf = base_recipe, xgb = base_recipe),
  models  = list(glm = glm_spec,   rf = rf_spec,      xgb = xgb_spec),
  cross   = FALSE
)
```
We are going to calculate a set of different performance metrics across CV folds.

- `pr_auc`: Area under the Precision-Recall curve. Sweeps the classification threshold from strict to lenient and plots precision against recall at each. Single-number summary of how well the model ranks septic patients above non-septic ones, without being inflated by the large true-negative pool. Baseline ≈ class prevalence (~0.02 here), so any value meaningfully above that is real signal. 

- `brier_class`: Mean squared error between predicted probabilities and the binary sepsis label 0/1 outcome. Measures **calibration + discrimination together** a model that says "80% sepsis risk" should be right 80% of the time. Lower is better; 0 is perfect, 0.25 is random guessing at 50/50. Unlike AUC metrics, Brier penalises *over-confident* probabilities

- `accuracy`: (TP+TN)/N. Due to imbalance this is included for familiarity and catching bugs! Reminder: Predicting "No Sepsis" for everyone scores ~98% here. 

- `sensitivity` (= recall, = true positive rate) TP/(TP+FN). Of patients who actually had sepsis, what fraction did we flag? The clinically critical metric for the positive class: missed sepsis (low sensitivity) means delayed antibiotics, which directly increases mortality.

- `specificity`: TN/(TN+FP). Of patients who did *not* have sepsis, what fraction did we correctly leave unflagged? The counterpart to sensitivity — high specificity means low false-alarm rate, which matters for **alarm fatigue** in the ICU. Sensitivity and specificity trade off as you move the threshold, so they're reported together.

Now we can run it, this will take a few minutes so we can use cache-ing again to avoid frustration!

```{r tune, cache = TRUE}
cls_metrics <- metric_set(pr_auc, brier_class, accuracy, sensitivity, specificity)

tuned <- wf_set |>
  workflow_map(
    fn        = "tune_grid",
    seed      = GLOBAL_SEED ,
    resamples = folds,
    grid      = 10,
    metrics   = cls_metrics,
    control   = control_grid(save_pred = TRUE, verbose = FALSE)
  )
```

### 5.5 Compare cross-validation results

```{r cv-comparison, fig.height = 3}
autoplot(tuned, metric = "pr_auc", select_best = TRUE) +
  labs(title = "Best CV PR-AUC per model family")
```

### Question 13

**Q13 (2 marks).** Which model family achieves the highest CV PR-AUC?
Is the gap between the best and worst model large relative to the standard errors?
What does that suggest about model selection here?

Looking at the plot, the logistic regression ranked first with a PR-AUC of ~0.76, random forest second at ~0.71, and XGBoost last at ~0.67. However, their confidence intervals are very wide and overlap substantially; random forest's interval runs from approximately 0.60 to 0.83, and XGBoost's from approximately 0.56 to 0.77. This means that the gap between the models is small relative to the uncertainty, so we cannot be confident that logistic regression is actually the best model. The ranking could easily change it the data is randomly split a different way.

```{r}
rank_results(tuned, rank_metric = "pr_auc", select_best = TRUE) |>
  filter(.metric %in% c("pr_auc", "brier_class", "accuracy", "sensitivity", "specificity")) |>
  select(wflow_id, .metric, mean, std_err, rank)
```

### Question 14

**Q14 (3 marks).** By summarising and comparing the metrics output by this code, describe how each model performed relative to one another.

Logistic regression performed the best overall, with the highest PR_AUC (0.761), highest accuracy (0.981), highest sensitivity (0.417), and lowest Brier score (0.0146). Random forest had the highest specificity of all three models at a perfect 1.000, meaning it produced no false positives at the default threshold, but its sensitivity (0.382) and PR-AUC (0.709) were slightly lower than the logistic regression. XGBoost was ranked third on PR-AUC (0.668) and accuracy (0.978), but had a sensitivity (0.400) slightly higher than the random forest. All three models had very high specificity (> 0.997) but low sensitivity (< 0.42), which reflects the class imbalance; predicting "No Sepsis" almost all the time will almost always be correct when sepsis only makes up approximately 3% of cases. The standard errors are large enough that the differences between the models are not conclusive, and all three would need to have different thresholds to actually be clinically useful to correctly identify sepsis cases.

---

## 6 Test-Set Evaluation

In the real-world we'd try to optimise these models further and/or try some different architectures but lets just take our best model and continue to test evaluation.

### 6.1 Finalise the best model

```{r finalise}
best_id <- rank_results(tuned, rank_metric = "pr_auc",
                        select_best = TRUE) |>
  filter(.metric == "pr_auc") |>
  slice_head(n = 1) |>
  pull(wflow_id)

final_wf <- extract_workflow(tuned, id = best_id) |>
  finalize_workflow(
    extract_workflow_set_result(tuned, id = best_id) |>
      select_best(metric = "pr_auc")
  )

split <- make_splits(train, test)
final_fit <- last_fit(final_wf, split = split, metrics = cls_metrics)
collect_metrics(final_fit)
```


```{r perf-curves, fig.width = 5.5, fig.height = 4}
preds <- collect_predictions(final_fit)

preds |>
  pr_curve(truth = SepsisLabel, .pred_Sepsis,
           event_level = "first") |>
  autoplot() + labs(title = "Precision–Recall curve (test set)")
```

### 6.3 Confusion matrix

### Question 15

**Q15 (1 marks).** How many times were true sepsis cases classified as non-sepsis in the test set? Is this more or less than the correct sepsis predictions?

True sepsis cases were classified as non-sepsis cases 18 times in the test set, which was 7 more times than it correctly identified sepsis cases (11).

```{r conf-matrix}
preds |>
  conf_mat(truth = SepsisLabel, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(title = "Confusion matrix (threshold = 0.5)")
```

---

## 7 Misclassification Analysis

### 7.1 Subgroup performance

Now let's investigate whether the model work equally well/badly across patient groups?

**Code deleted here due to it not working**

### Question 16

**Q16 (2 marks).** Is there a subgroup where the model performs noticeably worse than the overall PR-AUC?
Suggest one data-driven reason and one clinical reason that could explain the gap.

Looking at the subgroup PR-AUC plot, all subgroups performed far worse than the overall test PR-AUC of approximately 0.70. The 80+ age group had no bar at all, meaning the model could not produce a PR-AUC estimate for that group, likely because there were too few or even zero sepsis cases in that subgroup after the test split. Among groups with values, the 66-80 year old subgroup performed the best (~0.012) and the males (Sex = 1) performed the worst (~0.07), however they were all very similar and low.

A data-drive reason for this is that splitting the already small test set into subgroups leaves very few positive sepsis cases in each group, possibly none in certain groups, making the PR-AUC estimates impossible to calculate.

A clinical reason is that sepsis can present very differently across age and sex groups. For example, elderly patients are known to have a differnt sepsis path than younger or mild-aged adults (*Alhamyani et al., 2024*). As the model is trained on the full population, it may have not learned the sepsis patterns that are specific to each subgroup well enough to reliably identify cases within them.

Alhamyani AH, Alamri MS, Aljuaid NW, Aloubthani AH, Alzahrani S, Alghamdi AA, Lajdam AS, Alamoudi H, Alamoudi AA, Albulushi AM, AlQarni SN. Sepsis in Aging Populations: A Review of Risk Factors, Diagnosis, and Management. Cureus. 2024 Dec 2;16(12):e74973. doi:  (10.7759/cureus.74973)[https://pubmed.ncbi.nlm.nih.gov/39744263/]. PMID: 39744263; PMCID: PMC11691596.

!!!!!!!!!!!!!!! **fix citation**

### 7.2 Variable importance

```{r variable-importance, fig.height = 10}
parsnip_fit <- extract_fit_parsnip(final_fit)

importance_tbl <- if (grepl("^rf", best_id)) {
  imp <- parsnip_fit$fit$variable.importance
  tibble(variable = names(imp), importance = unname(imp))
} else if (grepl("^xgb", best_id)) {
  xgboost::xgb.importance(model = parsnip_fit$fit) |>
    as_tibble() |>
    transmute(variable = Feature, importance = Gain)
} else {
  broom::tidy(parsnip_fit) |>
    filter(term != "(Intercept)") |>
    transmute(variable = term, importance = abs(estimate))
}

importance_tbl |>
  slice_max(importance, n = 20) |>
  ggplot(aes(importance, fct_reorder(variable, importance))) +
  geom_col(fill = "steelblue") +
  labs(title = paste("Top variable importances: ", best_id),
       x = "Importance", y = NULL)
```

### Question 17

**Q17 (2 marks).** Compare the top 3 most important variables to the three predictors you identified in Q1.
Do they match? Write one sentence about any variable that surprised you.

The top three most important variables were lactate, MAP, and WBC, which did not match my Q2 predictions of Resp, SBP, and Temp. This makes sense looking back, as elevated lactate is one of the key indicators of septic shock in Sepsis-3 *(Singer et al., 2016)*, MAP looks at the entire state of circulation better than SBP alone, and high WBC is known as a classic sign of infection that would be present in sepsis cases as well. Resp and Temp had what appears to be xero importance, likely because they are elevated in many common ICU conditions and would not be specific to just sepsis. I was the most surprised that the respiratory rate had no importance, since it is used in both the SIRS  and qSOFA criteria.

Singer M, Deutschman CS, Seymour CW, Shankar-Hari M, Annane D, Bauer M, et al. The Third International Consensus Definitions for Sepsis and Septic Shock (Sepsis-3). JAMA. 2016 Feb 23;315(8):801–10. doi:[10.1001/jama.2016.0287](https://jamanetwork.com/journals/jama/fullarticle/2492881)

---

## 8 Discussion

### Question 18

**Q18 (3 marks).** In 3–5 sentences, address both of the following:

- What is the single biggest limitation of this analysis (consider the label definition, the static sampling approach, generalisability to other hospitals, etc.)?
- Would you recommend deploying a model trained this way in a real ICU? If yes, under what monitoring conditions? If no, what would need to change first?

The biggest limitation of this analysis is the static sampling approach, where a single random observation was selected from each patient's ICU stay. This removes all temporal information and prevents the model from capturing trends over time, such as vital signs and laboratory values getting worse, which would be signs that someone is developing sepsis or their sepsis is worsening.

I would not recommend deploying this model in a real ICU because the best model only achieved a sensitivity of 0.42, meaning that more than half of the true sepsis cases would be missed. Before being deployed, the model should be developed using the actual time-series data to see if that improves the classification. 

---

## Session Info

```{r session-info}
sessionInfo()
```

## Credits

This practical is adapted from earlier versions of the course's medical databases lab and updated to use the [PhysioNet/CinC Challenge 2019 sepsis dataset](https://physionet.org/content/challenge-2019/1.0.0/) (Reyna et al., *Crit Care Med* 2020) to use a more modern `tidymodels` workflow.
